Neural Co-Processors for Restoring Brain Function:\\Results from a Cortical Model of Grasping

### 2023-05-04 paper/article.tex
Final revision updates.
.tex may not match final text online, due to back-and-forth with editors.

\begin{figure}
    \centering
    \begin{subfigure}[c]{0.88\textwidth}
		\centering
		\includegraphics[width=\textwidth]{weill_arch.png}
		\caption{}
	\end{subfigure}
	\hfill
    \begin{subfigure}[c]{0.88\textwidth}
		\centering
		\includegraphics[width=\textwidth]{cpn_michaels_arch_labeled.eps}
		\caption{}
	\end{subfigure}
	\hfill
    \caption{{\bf Neural co-processor for restoring function after a brain injury}.
    An artificial neural network called the “Co-Processor Network” (CPN) is used to
    map input neural activity patterns in an area A to output stimulation patterns
    in same or other areas B in order to achieve a neural or behavioral goal using
    another ANN, an “Emulator Network” (EN) - see text for details. (a) The example
    here shows the CPN creating a new information processing pathway between prefrontal
    cortex and motor cortex, bypassing an intermediate area affected by brain injury
    (e.g., stroke). Adapted from \cite{rao.coproc}, (b) Our current study involves a
    simulated cortical grasping circuit. The CPN and EN both receive simulated brain
    recordings, which are created according to a recording model (see Methods).
    The CPN outputs stimulation parameters, which are applied to the simulated
    grasping circuit, according to a stimulation model. The EN models the relationship between
    the     stimulation, brain recordings, and external task.}
    \label{fig:arch}
\end{figure}

@@

\begin{figure}
	\centering
	\begin{subfigure}[c]{0.69\textwidth}
		\centering
		\includegraphics[width=\textwidth]{michaels.eps}
		\caption{}
	\end{subfigure}
	\hfill
	\begin{subfigure}[c]{0.30\textwidth}
		\centering
		\includegraphics[width=\textwidth]{mRNN_J.png}
		\caption{}
	\end{subfigure}
	\hfill
\caption{\textbf{Architecture of the Cortical Model for Grasping.} (a) Modular RNN (mRNN) used by Michaels et al.
        \cite{michaels.mrnn} to model cortical circuits involved in grasping objects.
        The emergent dynamics of the three modules correspond well to neural activity in
        primate cortical areas AIP, F5, M1, respectively, for the same task. Visual
        features of an object (derived from VGGNet) propagate forward through the
		network, conditioning the grasp on the object's size and shape. (b)
		Connectivity matrix $J$. Connected neuron pairs are indicated in black, though the
		actual network weights are floating point values. White indicates non-connections.
		Note the each module is fully connected (black squares along the diagonal) while connections to adjacent
        modules are sparse ($10\%$ of possible connections). All connections were trained using nonhuman primate
        (NHP) data to recapitulate the experimentally recorded muscle velocity trajectories for grasping different
        objects.}
\label{fig:michaels}
\end{figure}

The Michaels et al.\ cortical model implements a vision-to-grasp pipeline, from the visual
processing needed to move the hand to the appropriate position to shaping the hand for
grasping an object of a particular shape. The emergent dynamics of the model's modules, once trained,
correspond roughly to neural responses in the cortical areas AIP, F5, and M1 in a NHP subject's brain (Fig.~\ref{fig:michaels}a;
see \cite{michaels.mrnn} for details). For convenience, we will refer to the three
modules in the model using the cortical areas (AIP, F5, M1) they correspond to. For details on how this network model 
was trained, please see Supplementary Materials \ref{sup:michaelstraining}. For additional details on the task structure, see
Supplementary Materials \ref{sup:michaelstask}.

@@

On the other hand, if we lesion of a portion of the M1 module in the model,
we see a more complete loss of movement, affecting even the ability to reach for the grasp.
Finally, if we ``disconnect'' communication between the F5 and M1 modules,
we see a failure similar to an AIP lesion: the reaching movement is generally achieved, but
we see a disproportionate impact on forming the appropriate hand grasp. See Fig.
\ref{fig:lesion}a for examples of lesion impacts on velocities for individual muscles.
In Fig. \ref{fig:lesion}b we see the ratio of mean squared error (MSE) loss for hand
muscles to all muscles. Values greater than $1.0$ indicate that hand muscles are impacted more
than other muscles. Note that with the AIP and connection lesions, hand muscles show significantly
higher loss than overall loss. However, the M1 lesion shows the highest overall loss; see Table
\ref{tab:results} in Supplement materials section \ref{sup:results} for detailed lesion loss data.

\begin{figure}[h]
	\centering
	\begin{subfigure}[c]{0.62\textwidth}
	    \centering
	    \includegraphics[width=\textwidth]{lesion_trajs.pdf}
	    \caption{}
	\end{subfigure}
	\hfill
	\begin{subfigure}[c]{0.32\textwidth}
	    \centering
	    \includegraphics[width=\textwidth]{lesion_hand.pdf}
	    \caption{}
	\end{subfigure}
	\hfill
	\caption{\textbf{Simulating strokes by lesioning different parts of the model
             lead to differential impact on grasping behavior.}
             (a) Example muscle trajectories before (orange) and after (blue)
             simulated stroke due to AIP, F5-M1 and M1 lesions for a shoulder
             muscle (top row) and a hand muscle (bottom row). Lesions which prevent
             a forward propagation of visual information tend to have a larger effect
             on hand pose (example EIP muscle shown) than on the shoulder, which is more
             involved in the reach motion (example TMAJ muscle shown). Lesions of M1
             tend to cause a significant task loss on both.
             (b) Ratio of hand muscle mean squared error (MSE) losses vs. overall MSE
             losses, for different lesion types. The loss is measured relative to the
             model's movement trajectories prior to the lesion. Average of all trials,
             with $\pm1$ stdev shown. Higher ratios for AIP and connection lesions indicate those
             lesions cause a larger loss for hand movement than overall movement, though
             their overall loss is lower than for M1 (see Table \ref{tab:results}).}
	\label{fig:lesion}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\begin{subfigure}[c]{0.48\textwidth}
	    \centering
	    \includegraphics[width=\textwidth]{perturbe_single_tmaj.pdf}
	    \caption{}
	\end{subfigure}
	\hfill
	\begin{subfigure}[c]{0.48\textwidth}
	    \centering
	    \includegraphics[width=\textwidth]{perturbe_dist_tmaj.pdf}
	    \caption{}
	\end{subfigure}
	\hfill
	\begin{subfigure}[c]{0.48\textwidth}
	    \centering
	    \includegraphics[width=\textwidth]{perturbe_single_eip.pdf}
	    \caption{}
	\end{subfigure}
	\hfill
	\begin{subfigure}[c]{0.48\textwidth}
	    \centering
	    \includegraphics[width=\textwidth]{perturbe_dist_eip.pdf}
	    \caption{}
	\end{subfigure}
	\hfill
	\caption{\textbf{Perturbations of the simulated cortical network (i.e. due to stimulation)
	         have long-running effects on muscle activation.} Our results are consistent with
	         the fixed point analysis of the network in \cite{michaels.mrnn}, showing that the network
	         exhibits stability. It tends towards the original trajectory after perturbation.
	         Instantaneous random perturbation of a random group of 10 M1
	         neurons performed at time t=2s. (a, c) Single trial example for TMAJ, EIP muscles
	         respectively. (b, d) Distribution of effects across n=1000 samples on TMAJ, EIP muscles
	         respectively. Probability distribution shown to +/- 2 stdevs. Network exhibits similar
	         behavior when perturbance is applied at other time steps.}
	\label{fig:dynamics}
\end{figure}

@@

To make our stimulation model more realistic, rather than allowing stimulation to directly change the
output of single neurons, we simulated how stimulation may affect a network of neurons using a
model that incorporates aspects of both spatial and temporal smoothing. Our intent was not to create a
biophysical model of stimulation (our model does not arrange neurons in a volume to allow for such detailed
simulation). Instead, the stimulation model approximates the effects of \textit{in vivo} biological
stimulation (e.g., extracellular electrical stimulation) by diffusing the effects of stimulation across
many neurons and across time. Our EN must thus learn to approximate this stimulation model,
in addition to approximating the cortical dynamics of the brain and the mapping of those dynamics onto
the grasping outputs.\footnote{In Supplementary Materials section \ref{sup:passthrough}, we explore  
other stimulation models.}

@@

\begin{figure}[h]
	\begin{subfigure}[c]{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{stim_single.png}
		\caption{$\sigma=1.875$}
	\end{subfigure}
	\hfill
	\begin{subfigure}[c]{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{stim_single2.png}
		\caption{$\sigma=2.875$}
	\end{subfigure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=\textwidth]{stim_and_obs.png}
	\caption{\textbf{Example of stimulation and recording for a single trial.} Here, M1
	has been lesioned 50\%, indicated by hidden states of M1 neurons with a value of zero (bottom
	left panel, lowest part of the plot). We record neural activities (using the recording model) from only
	 the AIP and F5 modules, as explained in Section \ref{sec:experiments}. Stimulation is applied
	to M1 to drive the network's output. Stimulation is mostly constant until approximately t=250,
	when the hold signal was lifted and the reach began.}
	\label{fig:stim_and_obs}
\end{figure}

\subsubsection{Recording model}
To simulate recordings of our model neurons, we used a recording model that assumed that the model neurons
are laid out along a single spatial dimension with a given number of electrodes spread evenly apart.
We then applied Gaussian convolution, where the Gaussian kernel is centered at each electrode position.
Thus, the recording from each ``electrode'' is a Gaussian-weighted average of the activities of all model neurons
in the given module. Fig.~\ref{fig:stim_and_obs} (bottom right panel) shows an example of recordings obtained
using this recording model; note that the Gaussian averaging provides a potentially ambiguous view of the
activities of the underlying neurons due to the inherently destructive nature of averaging, making it
more challenging for the co-processor to interpret the neural activity and produce an appropriate
stimulation pattern. Our simulated recordings are similar in spirit, though not actually modeling,
local field potentials (LFPs) recorded extracellularly in biological neural tissue. In Supplementary Materials section
\ref{sup:observability}, we show that as we vary the observability of our ``simulated brain'' via this recording
model, the results remain largely the same, up to the point where the objects to be grasped can no longer be
distinguished from the neural recordings due to excessive averaging.

@@

\subsubsection{Simulating a non-stationary recording function, e.g., sensor drift}
Over time, implanted sensors may drift from true readings. We explored the co-processor's ability to adapt to
non-stationarity in the recording function. We performed an experiment where the recording function
includes an added bias term which changes over time, according to a random process. Each epoch, we
add a random value to each element of the bias term, drawn from a zero-mean Gaussian distribution, causing the
``recordings'' to drift over time. See Supplementary Materials section \ref{sup:drift} for further details.

@@

\subsubsection{Interleaved CPN/EN training, and adapting to non-stationarity}
\label{sec:interleaving}
Having defined the training procedures for the EN and CPN, we can define a training
protocol combining those two. We train the two in alternation, creating
a new EN each time we enter an EN training phase. We explored the possibility
of reusing an existing EN but found that retraining an EN was no more fast than training a new one.
Training a new EN also allows us to adapt to the brain's non-stationarity, which requires us to
determine when the current EN is no longer suitable for training the current CPN. 
Our algorithm retires the current EN and trains a new one when one of two conditions
is satisfied: (1) the EN's prediction error is sufficiently above some fraction of the CPN's task loss;
or (2) if CPN loss does not improve (or gets worse) over a sufficient number of recent training steps 
(this is a common stopping criterion in machine learning). Additional details can be found in the
Supplementary Materials section ~\ref{sup:encpninter}.

@@

\subsection{Stopping criteria}
In these experiments, we train the co-processor until one of the following two stopping criteria is
satisfied (in Supplementary Materials section \ref{sup:stopping_criteria}, we probe how these criteria
affect the results). The two criteria are:

\begin{itemize}
	\item The percent change in average task loss between two consecutive ranges of 500 epochs
	      is below a threshold, indicating minimal benefits of further training. The choice of
        threshold varies by experiment, due to the significant differences in setup. See Supplementary
        Material section \ref{sup:stopping_criteria} for further details.
	\item The training run has exceeded some number of epochs. This also varied by experiment.
\end{itemize}

@@

\subsection{Experiment 1: AIP 50\% lesion}
\begin{figure}[h]
\centering
\includegraphics[scale=1]{training_results_aip.png}
\caption{Co-Processor Performance Results: AIP 50\% Lesion. a. Percent loss recovery for the non-coadaptive
         experiment for the task and validation datasets. 0\% represents task loss equal to the loss for the
         lesioned cortical model, prior to any recovery. 100\% represents loss equal to that for the healthy
         cortical model. Negative initial values are due to the newly-initialized CPN causing
         worse performance than the lesioned model prior to recovery. b. Grasp separability metric ($S$) for the non-coadaptive
         experiment. Negative $S$ values indicate grasps are more similar between object types while 
         positive values indicate they are more distinct. c., d. show the performance of the neural co-processor for the
         co-adaptive experiment.}
\label{fig:results_aip}
\end{figure}

Our first experiment tested the ability of the co-processor to compensate for the loss of a significant
fraction (50\%) of the model neurons in the AIP module of the cortical model; this module is responsible for encoding
object shapes from visual inputs. Loss of AIP model neurons results in the cortical model, without further adaptation,
being unable to condition its grasp on the object shape, after largely succeeding in reaching toward the object.\footnote{
In Supplementary Materials section \ref{sup:results}, lesioned loss is lower in this experiment than in others: the subject
continues to reach successfully, but does not form the hand properly.} Likewise, the co-processor cannot observe sufficient
object shape encoding information due to the AIP lesion. As a result, as shown in Fig. \ref{fig:results_aip}a, significant
recovery is not possible. Fig. \ref{fig:results_aip}b shows that the co-processor never learns to separate object classes
($S$ value never approaches or exceeds $0.0$).

@@

\subsection{Experiment 2: M1 50\% lesion}
\begin{figure}[h]
\centering
\includegraphics[scale=1]{training_results_m1.png}
\caption{Co-Processor Performance Results: M1 50\% Lesion. a. Percent loss recovery
         for the non-coadaptive experiment for the task and validation datasets. See caption for Fig.~\ref{fig:results_aip}
         for interpretation of positive and negative values. b. Grasp separability metric (S) for the non-coadaptive experiment.
         See caption for Fig.~\ref{fig:results_aip} for interpretation of positive and negative values. c., d. show the
         performance of the neural co-processor for the co-adaptive experiment. 
         }
\label{fig:results_m1}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[scale=1]{training_results_con.png}
\caption{Co-Processor Performance Results: F5-M1 Connection Lesion. 
          a. Percent loss recovery for the non-coadaptive experiment for the task and validation datasets.
          See caption for Fig.~\ref{fig:results_aip} for interpretation of positive and negative values.
          b. Grasp separability metric (S) for the non-coadaptive experiment. See caption for Fig.~\ref{fig:results_aip} 
          or interpretation of positive and negative values. c., d. show the performance of the neural
          co-processor for the co-adaptive experiment.
         }
\label{fig:results_con}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[scale=1]{training_results_no_explore.png}
\caption{Co-Processor Performance Results: F5-M1 Connection Lesion: Use of exploratory
         and randomized examples in the training data set lead to stabilized
         and faster training. a., b. Without exploratory examples in the
         neighborhood of the CPN, we have unstable and slower training. In
         this case, the data set is composed of $10\%$ examples drawn from the
         CPN, and $90\%$ randomized examples (``No exploration'').
         c., d. Using only examples from the CPN (``Neither''; no exploratory or randomized examples) or $10\%$ CPN
         and $90\%$ exploratory examples (``No noise''; no randomized examples) also
         leads to unstable training. In c.\ and d., we truncate the experiments and use different
         axes to allow for clearer plots, but show the data from a.\ and b.\ for
         comparison.
         }
\label{fig:results_con_no_explore}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[scale=1]{training_results_recov.png}
\caption{Co-Processor Performance Results: Connection Lesion and Natural Recovery before Co-Processor Use.
         a. Percent loss recovery for the task and validation datasets. The horizontal dashed line is the task
         performance after both lesioning and natural recovery but before co-processor use. Initially the
         co-processor's stimulation causes task performance to be worse than the performance after natural
         recovery, but eventually the co-processor's performance exceeds this level. b. Grasp separability
         metric (S). Negative S values indicate grasps are more similar between object types while positive values
        indicate they are more distinct.
        }
\label{fig:results_recov}
\end{figure}

@@

\subsection{Experiment 5: Connection lesion with sensor drift}
Sensor-related non-stationarities are
often seen in real-world neural recording systems due to a variety of factors, from impedance
changes due to sensor movement to scar tissue formation. In our final experiment, we again lesioned the F5-M1
connections and allowed co-adaptation by the simulated cortical model but additionally, we allowed the sensor
readings to drift over time. This drift was modeled in the recording function, where the readings had a non-stationary
zero point across epochs. We added a bias term which we updated between epochs according to a random process
(see Supplementary Materials section~\ref{sup:drift} for details).

\begin{figure}[h]
\centering
\includegraphics[scale=1]{training_results_drift.png}
\caption{Co-Processor Performance Results: Connection Lesion with Sensor Drift. a. 
         Percent loss recovery for the task and validation datasets. 0\% represents task loss equal to the loss for
         the lesioned cortical model, prior to any recovery. 100\% represents
         loss equal to that for the healthy cortical model. 
         b. Grasp separability metric (S) which reflects the co-processor's ability
         to tailor grasps to specific objects. A negative value indicates the grasps are less
         differentiated between object shapes. Due to sensor drift, the co-processor
         is much slower to differentiate objects and the trend exhibits 
         higher noise compared to other experiments.}
\label{fig:results_drift}
\end{figure}

@@

\section{Supplementary Materials}
\subsection{Details of the simulated task}
\label{sup:michaelstask}

@@

\begin{figure}[h!]
\centering
\includegraphics[scale=0.7]{trial_timeline.png}
\caption{Timeline of a Single Trial}
\label{fig:trial_timeline}
\end{figure}

>>>>

\begin{figure}
    \centering
    \begin{subfigure}[c]{0.88\textwidth}
		\centering
		\includegraphics[width=\textwidth]{weill_arch.eps}
		\caption{}
	\end{subfigure}
	\hfill
    \begin{subfigure}[c]{0.88\textwidth}
		\centering
		\includegraphics[width=\textwidth]{cpn_michaels_arch_labeled.eps}
		\caption{}
	\end{subfigure}
	\hfill
    \caption{{\bf Neural co-processor for restoring function after a brain injury}.
    An artificial neural network called the “Co-Processor Network” (CPN) is used to
    map input neural activity patterns in an area A to output stimulation patterns
    in same or other areas B in order to achieve a neural or behavioral goal using
    another ANN, an “Emulator Network” (EN) - see text for details. (a) The example
    here shows the CPN creating a new information processing pathway between prefrontal
    cortex and motor cortex, bypassing an intermediate area affected by brain injury
    (e.g., stroke). Adapted from \cite{rao.coproc}, (b) Our current study involves a
    simulated cortical grasping circuit. The CPN and EN both receive simulated brain
    recordings, which are created according to a recording model (see Methods).
    The CPN outputs stimulation parameters, which are applied to the simulated
    grasping circuit, according to a stimulation model. The EN models the relationship between
    the     stimulation, brain recordings, and external task.}
    \label{fig:arch}
\end{figure}

@@

\begin{figure}
	\centering
	\begin{subfigure}[c]{0.69\textwidth}
		\centering
		\includegraphics[width=\textwidth]{michaels.eps}
		\caption{}
	\end{subfigure}
	\hfill
	\begin{subfigure}[c]{0.30\textwidth}
		\centering
		\includegraphics[width=\textwidth]{mRNN_J.eps}
		\caption{}
	\end{subfigure}
	\hfill
\caption{\textbf{Architecture of the Cortical Model for Grasping.} (a) Modular RNN (mRNN) used by Michaels et al.
        \cite{michaels.mrnn} to model cortical circuits involved in grasping objects.
        The emergent dynamics of the three modules correspond well to neural activity in
        primate cortical areas AIP, F5, M1, respectively, for the same task. Visual
        features of an object (derived from VGGNet) propagate forward through the
		network, conditioning the grasp on the object's size and shape. (b)
		Connectivity matrix $J$. Connected neuron pairs are indicated in black, though the
		actual network weights are floating point values. White indicates non-connections.
		Note the each module is fully connected (black squares along the diagonal) while connections to adjacent
        modules are sparse ($10\%$ of possible connections). All connections were trained using nonhuman primate
        (NHP) data to recapitulate the experimentally recorded muscle velocity trajectories for grasping different
        objects.}
\label{fig:michaels}
\end{figure}

The Michaels et al.\ cortical model implements a vision-to-grasp pipeline, from the visual
processing needed to move the hand to the appropriate position to shaping the hand for
grasping an object of a particular shape. The emergent dynamics of the model's modules, once trained,
correspond roughly to neural responses in the cortical areas AIP, F5, and M1 in a NHP subject's brain (Fig.~\ref{fig:michaels}a;
see \cite{michaels.mrnn} for details). For convenience, we will refer to the three
modules in the model using the cortical areas (AIP, F5, M1) they correspond to. For details on how this network model 
was trained, please see Appendix \ref{sup:michaelstraining}. For additional details on the task structure, see
Appendix \ref{sup:michaelstask}.

@@

On the other hand, if we lesion of a portion of the M1 module in the model,
we see a more complete loss of movement, affecting even the ability to reach for the grasp.
Finally, if we ``disconnect'' communication between the F5 and M1 modules,
we see a failure similar to an AIP lesion: the reaching movement is generally achieved, but
we see a disproportionate impact on forming the appropriate hand grasp. See Fig.
\ref{fig:lesion}a for examples of lesion impacts on velocities for individual muscles.
In Fig. \ref{fig:lesion}b we see the ratio of mean squared error (MSE) loss for hand
muscles to all muscles. Values greater than $1.0$ indicate that hand muscles are impacted more
than other muscles. Note that with the AIP and connection lesions, hand muscles show significantly
higher loss than overall loss. However, the M1 lesion shows the highest overall loss; see Table
\ref{tab:results} in Appendix \ref{sup:results} for detailed lesion loss data.

\begin{figure}[h]
	\centering
	\begin{subfigure}[c]{0.62\textwidth}
	    \centering
	    \includegraphics[width=\textwidth]{lesion_trajs.eps}
	    \caption{}
	\end{subfigure}
	\hfill
	\begin{subfigure}[c]{0.32\textwidth}
	    \centering
	    \includegraphics[width=\textwidth]{lesion_hand.eps}
	    \caption{}
	\end{subfigure}
	\hfill
	\caption{\textbf{Simulating strokes by lesioning different parts of the model
             lead to differential impact on grasping behavior.}
             (a) Example muscle trajectories before (orange) and after (blue)
             simulated stroke due to AIP, F5-M1 and M1 lesions for a shoulder
             muscle (top row) and a hand muscle (bottom row). Lesions which prevent
             a forward propagation of visual information tend to have a larger effect
             on hand pose (example EIP muscle shown) than on the shoulder, which is more
             involved in the reach motion (example TMAJ muscle shown). Lesions of M1
             tend to cause a significant task loss on both.
             (b) Ratio of hand muscle mean squared error (MSE) losses vs. overall MSE
             losses, for different lesion types. The loss is measured relative to the
             model's movement trajectories prior to the lesion. Average of all trials,
             with $\pm1$ stdev shown. Higher ratios for AIP and connection lesions indicate those
             lesions cause a larger loss for hand movement than overall movement, though
             their overall loss is lower than for M1 (see Table \ref{tab:results}).}
	\label{fig:lesion}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\begin{subfigure}[c]{0.48\textwidth}
	    \centering
	    \includegraphics[width=\textwidth]{perturbe_single_tmaj.eps}
	    \caption{}
	\end{subfigure}
	\hfill
	\begin{subfigure}[c]{0.48\textwidth}
	    \centering
	    \includegraphics[width=\textwidth]{perturbe_dist_tmaj.eps}
	    \caption{}
	\end{subfigure}
	\hfill
	\begin{subfigure}[c]{0.48\textwidth}
	    \centering
	    \includegraphics[width=\textwidth]{perturbe_single_eip.eps}
	    \caption{}
	\end{subfigure}
	\hfill
	\begin{subfigure}[c]{0.48\textwidth}
	    \centering
	    \includegraphics[width=\textwidth]{perturbe_dist_eip.eps}
	    \caption{}
	\end{subfigure}
	\hfill
	\caption{\textbf{Perturbations of the simulated cortical network (i.e. due to stimulation)
	         have long-running effects on muscle activation.} Our results are consistent with
	         the fixed point analysis of the network in \cite{michaels.mrnn}, showing that the network
	         exhibits stability. It tends towards the original trajectory after perturbation.
	         Instantaneous random perturbation of a random group of 10 M1
	         neurons performed at time t=2s. (a, c) Single trial example for TMAJ, EIP muscles
	         respectively. (b, d) Distribution of effects across n=1000 samples on TMAJ, EIP muscles
	         respectively. Probability distribution shown to +/- 2 stdevs. Network exhibits similar
	         behavior when perturbance is applied at other time steps.}
	\label{fig:dynamics}
\end{figure}

@@

To make our stimulation model more realistic, rather than allowing stimulation to directly change the
output of single neurons, we simulated how stimulation may affect a network of neurons using a
model that incorporates aspects of both spatial and temporal smoothing. Our intent was not to create a
biophysical model of stimulation (our model does not arrange neurons in a volume to allow for such detailed
simulation). Instead, the stimulation model approximates the effects of \textit{in vivo} biological
stimulation (e.g., extracellular electrical stimulation) by diffusing the effects of stimulation across
many neurons and across time. Our EN must thus learn to approximate this stimulation model,
in addition to approximating the cortical dynamics of the brain and the mapping of those dynamics onto
the grasping outputs.\footnote{In Appendix \ref{sup:passthrough}, we explore
other stimulation models.}

@@

\begin{figure}[h]
	\begin{subfigure}[c]{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{stim_single1.eps}
		\caption{$\sigma=1.875$}
	\end{subfigure}
	\hfill
	\begin{subfigure}[c]{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{stim_single2.eps}
		\caption{$\sigma=2.875$}
	\end{subfigure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=\textwidth]{stim_and_obs.eps}
	\caption{\textbf{Example of stimulation and recording for a single trial.} Here, M1
	has been lesioned 50\%, indicated by hidden states of M1 neurons with a value of zero (bottom
	left panel, lowest part of the plot). We record neural activities (using the recording model) from only
	 the AIP and F5 modules, as explained in Section \ref{sec:experiments}. Stimulation is applied
	to M1 to drive the network's output. Stimulation is mostly constant until approximately t=250,
	when the hold signal was lifted and the reach began.}
	\label{fig:stim_and_obs}
\end{figure}

\subsubsection{Recording model}
To simulate recordings of our model neurons, we used a recording model that assumed that the model neurons
are laid out along a single spatial dimension with a given number of electrodes spread evenly apart.
We then applied Gaussian convolution, where the Gaussian kernel is centered at each electrode position.
Thus, the recording from each ``electrode'' is a Gaussian-weighted average of the activities of all model neurons
in the given module. Fig.~\ref{fig:stim_and_obs} (bottom right panel) shows an example of recordings obtained
using this recording model; note that the Gaussian averaging provides a potentially ambiguous view of the
activities of the underlying neurons due to the inherently destructive nature of averaging, making it
more challenging for the co-processor to interpret the neural activity and produce an appropriate
stimulation pattern. Our simulated recordings are similar in spirit, though not actually modeling,
local field potentials (LFPs) recorded extracellularly in biological neural tissue. In Appendix
\ref{sup:observability}, we show that as we vary the observability of our ``simulated brain'' via this recording
model, the results remain largely the same, up to the point where the objects to be grasped can no longer be
distinguished from the neural recordings due to excessive averaging.

@@

\subsubsection{Simulating a non-stationary recording function, e.g., sensor drift}
Over time, implanted sensors may drift from true readings. We explored the co-processor's ability to adapt to
non-stationarity in the recording function. We performed an experiment where the recording function
includes an added bias term which changes over time, according to a random process. Each epoch, we
add a random value to each element of the bias term, drawn from a zero-mean Gaussian distribution, causing the
``recordings'' to drift over time. See Appendix \ref{sup:drift} for further details.

@@

\subsubsection{Interleaved CPN/EN training, and adapting to non-stationarity}
\label{sec:interleaving}
Having defined the training procedures for the EN and CPN, we can define a training
protocol combining those two. We train the two in alternation, creating
a new EN each time we enter an EN training phase. We explored the possibility
of reusing an existing EN but found that retraining an EN was no more fast than training a new one.
Training a new EN also allows us to adapt to the brain's non-stationarity, which requires us to
determine when the current EN is no longer suitable for training the current CPN. 
Our algorithm retires the current EN and trains a new one when one of two conditions
is satisfied: (1) the EN's prediction error is sufficiently above some fraction of the CPN's task loss;
or (2) if CPN loss does not improve (or gets worse) over a sufficient number of recent training steps 
(this is a common stopping criterion in machine learning). Additional details can be found in the
Appendix ~\ref{sup:encpninter}.

@@

\subsection{Stopping criteria}
In these experiments, we train the co-processor until one of the following two stopping criteria is
satisfied (in Appendix \ref{sup:stopping_criteria}, we probe how these criteria
affect the results). The two criteria are:

\begin{itemize}
	\item The percent change in average task loss between two consecutive ranges of 500 epochs
	      is below a threshold, indicating minimal benefits of further training. The choice of
              threshold varies by experiment, due to the significant differences in setup. See Appendix
              \ref{sup:stopping_criteria} for further details.
	\item The training run has exceeded some number of epochs. This also varied by experiment.
\end{itemize}

@@

\subsection{Experiment 1: AIP 50\% lesion}
\begin{figure}[h]
\centering
\includegraphics[scale=1]{training_results_aip.eps}
\caption{Co-Processor Performance Results: AIP 50\% Lesion. a. Percent loss recovery for the non-coadaptive
         experiment for the task and validation datasets. 0\% represents task loss equal to the loss for the
         lesioned cortical model, prior to any recovery. 100\% represents loss equal to that for the healthy
         cortical model. Negative initial values are due to the newly-initialized CPN causing
         worse performance than the lesioned model prior to recovery. b. Grasp separability metric ($S$) for the non-coadaptive
         experiment. Negative $S$ values indicate grasps are more similar between object types while 
         positive values indicate they are more distinct. c., d. show the performance of the neural co-processor for the
         co-adaptive experiment.}
\label{fig:results_aip}
\end{figure}

Our first experiment tested the ability of the co-processor to compensate for the loss of a significant
fraction (50\%) of the model neurons in the AIP module of the cortical model; this module is responsible for encoding
object shapes from visual inputs. Loss of AIP model neurons results in the cortical model, without further adaptation,
being unable to condition its grasp on the object shape, after largely succeeding in reaching toward the object.\footnote{
In Appendix \ref{sup:results}, lesioned loss is lower in this experiment than in others: the subject
continues to reach successfully, but does not form the hand properly.} Likewise, the co-processor cannot observe sufficient
object shape encoding information due to the AIP lesion. As a result, as shown in Fig. \ref{fig:results_aip}a, significant
recovery is not possible. Fig. \ref{fig:results_aip}b shows that the co-processor never learns to separate object classes
($S$ value never approaches or exceeds $0.0$).

@@

\subsection{Experiment 2: M1 50\% lesion}
\begin{figure}[h]
\centering
\includegraphics[scale=1]{training_results_m1.eps}
\caption{Co-Processor Performance Results: M1 50\% Lesion. a. Percent loss recovery
         for the non-coadaptive experiment for the task and validation datasets. See caption for Fig.~\ref{fig:results_aip}
         for interpretation of positive and negative values. b. Grasp separability metric (S) for the non-coadaptive experiment.
         See caption for Fig.~\ref{fig:results_aip} for interpretation of positive and negative values. c., d. show the
         performance of the neural co-processor for the co-adaptive experiment. 
         }
\label{fig:results_m1}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[scale=1]{training_results_con.eps}
\caption{Co-Processor Performance Results: F5-M1 Connection Lesion. 
          a. Percent loss recovery for the non-coadaptive experiment for the task and validation datasets.
          See caption for Fig.~\ref{fig:results_aip} for interpretation of positive and negative values.
          b. Grasp separability metric (S) for the non-coadaptive experiment. See caption for Fig.~\ref{fig:results_aip} 
          or interpretation of positive and negative values. c., d. show the performance of the neural
          co-processor for the co-adaptive experiment.
         }
\label{fig:results_con}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[scale=1]{training_results_no_explore.eps}
\caption{Co-Processor Performance Results: F5-M1 Connection Lesion: Use of exploratory
         and randomized examples in the training data set lead to stabilized
         and faster training. a., b. Without exploratory examples in the
         neighborhood of the CPN, we have unstable and slower training. In
         this case, the data set is composed of $10\%$ examples drawn from the
         CPN, and $90\%$ randomized examples (``No exploration'').
         c., d. Using only examples from the CPN (``Neither''; no exploratory or randomized examples) or $10\%$ CPN
         and $90\%$ exploratory examples (``No noise''; no randomized examples) also
         leads to unstable training. In c.\ and d., we truncate the experiments and use different
         axes to allow for clearer plots, but show the data from a.\ and b.\ for
         comparison.
         }
\label{fig:results_con_no_explore}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[scale=1]{training_results_recov.eps}
\caption{Co-Processor Performance Results: Connection Lesion and Natural Recovery before Co-Processor Use.
         a. Percent loss recovery for the task and validation datasets. The horizontal dashed line is the task
         performance after both lesioning and natural recovery but before co-processor use. Initially the
         co-processor's stimulation causes task performance to be worse than the performance after natural
         recovery, but eventually the co-processor's performance exceeds this level. b. Grasp separability
         metric (S). Negative S values indicate grasps are more similar between object types while positive values
        indicate they are more distinct.
        }
\label{fig:results_recov}
\end{figure}

@@

\subsection{Experiment 5: Connection lesion with sensor drift}
Sensor-related non-stationarities are
often seen in real-world neural recording systems due to a variety of factors, from impedance
changes due to sensor movement to scar tissue formation. In our final experiment, we again lesioned the F5-M1
connections and allowed co-adaptation by the simulated cortical model but additionally, we allowed the sensor
readings to drift over time. This drift was modeled in the recording function, where the readings had a non-stationary
zero point across epochs. We added a bias term which we updated between epochs according to a random process
(see Appendix \ref{sup:drift} for details).

\begin{figure}[h]
\centering
\includegraphics[scale=1]{training_results_drift.eps}
\caption{Co-Processor Performance Results: Connection Lesion with Sensor Drift. a.
         Percent loss recovery for the task and validation datasets. 0\% represents task loss equal to the loss for
         the lesioned cortical model, prior to any recovery. 100\% represents
         loss equal to that for the healthy cortical model.
         b. Grasp separability metric (S) which reflects the co-processor's ability
         to tailor grasps to specific objects. A negative value indicates the grasps are less
         differentiated between object shapes. Due to sensor drift, the co-processor
         is much slower to differentiate objects and the trend exhibits
         higher noise compared to other experiments.}
\label{fig:results_drift}
\end{figure}

@@

\section{Appendix}
\subsection{Details of the simulated task}
\label{sup:michaelstask}

@@

\begin{figure}[h!]
\centering
\includegraphics[scale=0.7]{trial_timeline.eps}
\caption{Timeline of a Single Trial}
\label{fig:trial_timeline}
\end{figure}
